The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-06-01 paper.tex
integrate problems and research gaps

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
\todo{Autorenreihenfolge festlegen}
% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
% TODO: overleaf
% Two or more authors:
\begin{aug}
\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e4]{laura.penne@hs-mittweida.de}}
\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e5]{fritzsc2@hs-mittweida.de}}
\author[B]{\inits{N.}\fnms{Niklas} \snm{de Sousa Norte}\ead[label=e6]{niklas.desousanorte@hs-mittweida.de}}
\author[B]{\inits{A.}\fnms{Andrea} \snm{Ferencová}\ead[label=e7]{andrea.ferencova@hs-mittweida.de}}
\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e8]{andy.ludwig@hs-mittweida.de}}% ist jetzt beim Fraunhofer, möchte aber weiter so gelistet werden
\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e9]{kreuzer@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
%\author[B]{\inits{}\fnms{} \snm{}\ead[label=e9]{}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6,e7,e8,e9,e10}}
\end{aug}

@@

Data from historical and prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allows for a comprehensive understanding of deceased individuals from different time periods, ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
For example, anthropological methods can be used to determine individual properties (phenotypes) such as their sex, age, height and ancestry and further individualising traits.

Digitalisation offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
Moreover it allows for linking data of different study samples across geographical boundaries.
Data sets consolidated this way and formalised through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.

Apart from contextual information, most informative clues can be found directly on the bone.
Anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.

Applying the same principles used for mapping the Earth’s surface \enquote{terrain}, it is possible to create a detailed map of the human body containing visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}. ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, ensure their easy retrieval and furthermore serve as a basis for the objective examination, for instance in the form of measurements.

ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures. In its current version, the ontology  refers to standardised normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors. ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analysing 3D-models of human skeletal remains.

\section{Related Work}
There are various nomenclatures for the description of anatomical landmarks.
These allow a uniform classification of anatomical features and structures:

The \emph{Terminologia anatomica}~\citep{ta2} (TA) is a hierarchy of anatomical landmarks of the entire human body.
From its history, the Latin and Greek designations have emerged to allow standardization.
It is advantageous that these languages no longer show progress and are increasingly losing political power.
In TA, the order of structures follows the natural anatomy.
An anatomical landmark is identified by an individual identification number, the anatomical term, and the English term.
The identification number is critical because certain landmarks are redundant.
For example, there is the \emph{processus zygomaticus} at the \emph{os frontale} and the \emph{os temporale}.
Synonyms, like \emph{zygomatic bone} and \emph{malar bone}, are also given.
Terms included in the \emph{other} category include eponyms, which are concepts named after people, places or things.
For example, \emph{ossa digitorum pedis} (engl. \emph{phalanges of the foot}) are the small bones of the toes and are also denoted as \emph{ossa phalangea pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
% chapter 2: Ossa (Bones)

The \emph{Foundational Model of Anatomy} (FMA) ontology~\citep{fma} represents the physical organization of human anatomy.
%It allows the knowledge it contains to be represented in a human-understandable and machine-interpretable way.
However, it does not include Greek and Latin labels, which are frequently used by medical professionals.
Due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
This has a negative impact on the comprehensibility for users.

Furthermore, in both nomenclatures there is an inconsistent use of singular and plural forms, gaps in lateralized landmarks and numbered series of certain bones (e.g. costae).
In anatomy, it is also important to distinguish between the medical and anthropological viewpoints.
For example, relevant anthropological aspects such as the degree of wear of bones are not considered.
Furthermore, the articular surfaces of the individual bones are particularly important in anthropology.
Not all articular surfaces (e.g., on the ossa metatarsalia) are defined in the nomenclatures, resulting in further gaps.
In some cases, contradictions are also found in the definitions of individual landmarks.
For example, the corpus ossis pubis is placed elsewhere in some sources than in others.
Another problem is that the individual services do not synchronize and do not offer enough synonyms. % which services?
In addition, the anatomical landmarks are not visually represented and are only selectively labeled in known anatomy atlases.
However, since landmarks are three-dimensional structures, it is essential to provide a marker over the entire landmark and delineate it accurately.
The delineations must also be clearly described textually.
However, there are no standardized definitions in anatomy so far and also in the literature the information is sparse.
For example, in some sources only punctual labels are presented on a figure.
A similar situation exists with osteometric landmarks.
Standardized ones are found only on the cranium and mandible.
However, since there are measurement distances on all bones, the respective start and end points must also be defined.
These are stated differently in the literature and have a high number of synonyms.

For these reasons, it is necessary to develop uniform, textual, visual and detailed descriptive definitions for the individual landmarks, as well as to represent them in a new and clear way in an ontology.
The ontology should be easy to understand for the user.

@@

There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature - general or subject-specific - can address adequately so far.

The primary issue lies in the absence of standardisation. While both the TA and FMA have been proposed as a means of standardising terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardisation in actual practice~\citep{frequencyta,doestamatter,athighlights}. 
Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.  
Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas. As a consequence there are at times various translations for the same anatomical terms. Yet, knowledge of Latin is generally declining. With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes.
Moreover, in textual descriptions latin and the native language are commonly used interchangeably for stylistic purposes. For instance <<{{Quelle}}> uses the German terms of the bones in the explanation of sutures. 
Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example,  there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.

If not historically evolved, anatomical terms usually correspond to a description of the location, affiliation or function of the respective concept or structure. However, this kind of naming variable, non-intuitive and therefore inconsistent. 
An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in {{Quelle}} is described as saddle-shaped although it is not called ‚sella‘, the Latin name for saddle. It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.  
The Cuspis dentis, the tooth cusp, can by shape also be described as a protuberance. The Protuberantia mentale meanwhile is referred to as Eminentia mentale in a widely-used anthropological method for sex determination. 
As can be seen, the incentive for synonyms is high. However, the current resources fail to provide comprehensive lists of synonyms in use.  This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the „preferred name„ External acoustic aperture and „non-English equivalent“  Porus acusticus externus (FMA ID: 61301) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym. However, it is not the same, rather the Porus is - as its English name indicates - the opening to the meatus.   
Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong. When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures. For instance, the position of the Corpus ossis pubis varies across different sources.

In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.
Another main issue  is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field. For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g. on the Ossa carpi or Ossa metatarsi) are defined in the TA or FMA, resulting in further gaps. 
A similar situation exists with respect to osteometric measurements. Established and to some extent standardised osteometric measurement points are only found on the cranium and mandible.  However, since there are measurement distances on all bones, the respective start and end points must also be defined. Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants.
Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures, which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general anatomy . A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other ‘articular eminence’. Often exchange between these specific fields and general anatomy hardly takes place. It stands to reason that interdisciplinary coordination between the individual disciplines and general anatomy hardly takes place.

>>>>

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
\todo{Autorenreihenfolge festlegen}
% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
% TODO: overleaf
% Two or more authors:
\begin{aug}
\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}
\thanks{Equal contribution.}%how to show this for the first 3?
}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e4]{laura.penne@hs-mittweida.de}}
\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e5]{fritzsc2@hs-mittweida.de}}
\author[B]{\inits{N.}\fnms{Niklas} \snm{de Sousa Norte}\ead[label=e6]{niklas.desousanorte@hs-mittweida.de}}
\author[B]{\inits{A.}\fnms{Andrea} \snm{Ferencová}\ead[label=e7]{andrea.ferencova@hs-mittweida.de}}
\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e8]{andy.ludwig@hs-mittweida.de}}% ist jetzt beim Fraunhofer, möchte aber weiter so gelistet werden
\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e9]{kreuzer@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
%\author[B]{\inits{}\fnms{} \snm{}\ead[label=e9]{}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6,e7,e8,e9,e10}}
\end{aug}

@@

\section{Related Work}

@@

There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately so far.

The primary issue lies in the absence of standardisation.
While both the TA and FMA have been proposed as a means of standardising terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardisation in actual practice~\citep{frequencyta,doestamatter,athighlights}.
Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
Yet, knowledge of Latin is generally declining.
With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
Moreover, in textual descriptions latin and the native language are commonly used interchangeably for stylistic purposes.
For instance, \citet{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example,  there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.

If not historically evolved, anatomical terms usually correspond to a description of the location, affiliation or function of the respective concept or structure~\citep{reforminganatomical}.
However, this kind of naming variable, non-intuitive and therefore inconsistent.
An example, showing the variability in illustrative descriptions are the Tuberculum articulare of the Processus zygomaticus, which in \citet{anatomie} is described as saddle-shaped although it is not called \emph{sella}, the Latin name for saddle.
It is moreover worth pointing out that not every anatomical structures contributing to an articulation in some way, automatically carry the term articular with them, such as the Caput mandibulae, which connects the Mandibula to the rest of the Cranium.
The \emph{Cuspis dentis}, the tooth cusp, can by shape also be described as a protuberance~\citep{pocketatlas,anatomylexicon2008}.
The Protuberantia mentale meanwhile is referred to as Eminentia mentale in a widely-used anthropological method for sex determination~\citep{datacollection}.
As can be seen, the incentive for synonyms is high.
However, the current resources fail to provide comprehensive lists of synonyms in use.
 This in turn may lead to misattribution, as is the case for example, in the FMA the entry with the \emph{preferred name} \enquote{External acoustic aperture} and \emph{non-English equivalent} \enquote{Porus acusticus externus} (FMA ID: 61301) lists external acoustic meatus (Meatus acusticus externus in Latin) as a synonym.
However, it is not the same, rather the Porus is---as its English name indicates---the opening to the meatus~\citep{anatomylexicon}.
Furthermore, terms oftentimes lack clear affiliation to the bone to which they belong.
When considering the whole skeleton, these terms consequently occur multiple times, causing ambiguity.
Occasionally, there are discrepancies and inconsistencies found in the definitions of individual anatomical structures.
For instance, the position of the Corpus ossis pubis varies across different sources~\citep{anatomylexicon,prometheus,allgemeineanatomie,dualereiheanatomie,anatomiedesmenschen}.

In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.%add source for that? 11 and 12 in table but no paper.
Another main issue  is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialised field~\citep{fma}.
For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g. on the Ossa carpi or Ossa metatarsi) are defined in the TA or FMA, resulting in further gaps.
A similar situation exists with respect to osteometric measurements.
Established and to some extent standardised osteometric measurement points are only found on the cranium and mandible~\citep{wesenanthropologie}.
 However, since there are measurement distances on all bones, the respective start and end points must also be defined.
Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants~\citep{anatomycontribution}.
Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures~\citep{graysanatomy},
which to the greater part are not found in the FMA or the TA that primarily aim for the standardisation of general anatomy~\citep{fma}.
A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other \enquote{articular eminence}~\citep{dentalanthropology}.
Often exchange between these specific fields and general anatomy hardly takes place.
It stands to reason that interdisciplinary coordination between the individual disciplines and general anatomy hardly takes place.
